---
title: "Data Analysis Assignment #2 (75 points total)"
author: "De, Abhijit"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

### Instructions

R markdown is a plain-text file format for integrating text and R code, and creating transparent, reproducible and interactive reports. An R markdown file (.Rmd) contains metadata, markdown and R code "chunks", and can be "knit" into numerous output types. Answer the test questions by adding R code to the fenced code areas below each item. There are questions that require a written answer that also need to be answered. Enter your comments in the space provided as shown below:

***Answer: (Enter your answer here.)*** 

Once completed, you will "knit" and submit the resulting .html document and the .Rmd file. The .html will present the output of your R code and your written answers, but your R code will not appear.  Your R code will appear in the .Rmd file. The resulting .html document will be graded and a feedback report returned with comments.  Points assigned to each item appear in the template.

**Before proceeding, look to the top of the .Rmd for the (YAML) metadata block, where the *title*, *author* and *output* are given. Please change *author* to include your name, with the format 'lastName, firstName.'**

If you encounter issues with knitting the .html, please send an email via Canvas to your TA.

Each code chunk is delineated by six (6) backticks; three (3) at the start and three (3) at the end. After the opening ticks, arguments are passed to the code chunk and in curly brackets. **Please do not add or remove backticks, or modify the arguments or values inside the curly brackets**. An example code chunk is included here: 

```{r exampleCodeChunk, eval = FALSE, echo = TRUE}
# Comments are included in each code chunk, simply as prompts

#...R code placed here

#...R code placed here

```

R code only needs to be added inside the code chunks for each assignment item. However, there are questions that follow many assignment items. Enter your answers in the space provided. An example showing how to use the template and respond to a question follows.

-----

**Example Problem with Solution:**

Use *rbinom()* to generate two random samples of size 10,000 from the binomial distribution. For the first sample, use p = 0.45 and n = 10. For the second sample, use p = 0.55 and n = 10. Convert the sample frequencies to sample proportions and compute the mean number of successes for each sample. Present these statistics.

```{r Example, eval = TRUE, echo = TRUE}

set.seed(123)
sample.one <- table(rbinom(10000, 10, 0.45)) / 10000
sample.two <- table(rbinom(10000, 10, 0.55)) / 10000

successes <- seq(0, 10)

round(sum(sample.one*successes), digits = 1) # [1] 4.5
round(sum(sample.two*successes), digits = 1) # [1] 5.5
```

**Question: How do the simulated expectations compare to calculated binomial expectations?**

***Answer:  The calculated binomial expectations are 10(0.45) = 4.5 and 10(0.55) = 5.5.  After rounding the simulated results, the same values are obtained.***

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(dplyr)
library(flux)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rockchalk)
library(tidyverse)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.
getwd()
setwd("C:/Users/JM933JS/Downloads/MSDS/Stat/r_code/MSDS401")

mydata <- read.csv("mydata.csv", sep = ",", stringsAsFactors = TRUE)
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)
summary(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

***#### Section 1: (5 points) ####***

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a, fig.width = 12}

#plot histogram and QQ plot
par(mfrow = c(1,2))
hist(mydata$RATIO, main = "Histogram of Ratio", xlab = "RATIO", col ="purple")
qqnorm(mydata$RATIO, main = "Q-Q Plot of Ratio",col = "red")
qqline(mydata$RATIO, col = "darkred")
par(mfrow = c(1,1))

#skewness and kurtosis using rockchalk
skew_rockchalk <- rockchalk::skewness(mydata$RATIO)
kurt_rockchalk_sub <- rockchalk::kurtosis(mydata$RATIO)
kurt_rockchalk <- rockchalk::kurtosis(mydata$RATIO,excess = FALSE)

#skewness and kurtosis using momnents
skew_moments <- moments::skewness(mydata$RATIO)
kurt_moments <- moments::kurtosis(mydata$RATIO)

cat("Skewness using moments : ", skew_moments, "\n")
cat("Skewness using rockchalk : ", skew_rockchalk, "\n")
cat("Kurtosis using moments : ", kurt_moments, "\n")
cat("Kurtosis using rockchalk after subtracting 3 : ", kurt_rockchalk_sub, "\n")
cat("Kurtosis using rockchalk without subtracting : ", kurt_rockchalk)

```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b, fig.width = 12, fig.height = 8}

mydata$L_RATIO <- log10(mydata$RATIO)

# histogram and QQ plot using L_RATIO
par(mfrow = c(1,2))

hist(mydata$L_RATIO, main = "Histogram of Log 10 of RATIO", xlab = "Log 10 of Ratio", col = "green")
qqnorm(mydata$L_RATIO, main = "Q-Q Plot of Log 10 of Ratio",col = "red")
qqline(mydata$L_RATIO, col = "darkred")

par(mfrow = c(1,1))

#Calculate the skewness and kurtosis
skew_lratio <- rockchalk::skewness(mydata$L_RATIO)
kurt_lratio <- rockchalk::kurtosis(mydata$L_RATIO,excess = FALSE)
cat("Skewness using moments : ", skew_lratio, "\n")
cat("Kurtosis using rockchalk without subtracting : ", kurt_lratio, "\n")

#Create a boxplot of L_RATIO differentiated by CLASS
boxplot(mydata$L_RATIO ~ mydata$CLASS,main = "Log 10 of RATIO by Class", col = "purple",xlab = "Class")


```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}

bartlett.test(mydata$RATIO ~ mydata$CLASS)
bartlett.test(mydata$L_RATIO ~ mydata$CLASS)

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: We see that L_Ratio exhibits better conformance to a normal distribution with homogeneous variances across age classes. For Ratio, we see in the QQ-Plot that the distribution tends to deviate from normality and this is confirmed by the skewness and kurtosis. We also see that once we transform the variable to a log10 form, the kurtosis and skewness resemble more to a normal distribution. This is also confirmed by the histogram and QQ-Plot. We also run the Bartlett test to test for the homogeneity of variance for both RATIO and L_Ratio but fail to reject the null hypothesis for L_Ratio. Therefore, we assume homogeneity across classes for L_Ratio.  ***


***#### Section 2 (10 points) ####***

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}

AOV1 <- aov(L_RATIO~CLASS+SEX+CLASS*SEX, mydata)
AOV2 <- aov(L_RATIO~CLASS+SEX, mydata)
summary(AOV1)
summary(AOV2)

```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: CLASS:SEX has a non-significant p-value when predicting L_RATIO compared to CLASS and SEX; the interaction term has no effect and therefore was not significant. The other aforementioned two variables, CLASS and SEX, are significant when predicting L_RATIO.  ***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}

TukeyHSD(AOV2,conf.level = 0.95)

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: There seems to be no significant difference between classes A1 and A2, in accordance to the high p value = 0.6919456. However, there are markedly significant differences between all other classes; all other class comparisons surpass threshold p-value of 0.05. Secondly, based on the high p-value of 0.9412689 between male and female abalones, we can determine there is no significant difference between male and female abalones. Threshold p-value of 0.05 would not allow us to reject this null hypothesis between male and female; thus, it might be best to combine them to a single category labelled as "adults". ***


***####  Section 3: (10 points) ####***

(3)(a1) Here, we will combine "M" and "F" into a new level, "ADULT". The code for doing this is given to you. For (3)(a1), all you need to do is execute the code as given.

```{r Part_3a1}
# Here, we show how to define the new variable TYPE using only base R functions:

mydata$TYPE <- factor(ifelse(mydata$SEX == "I", "I", "ADULT"))
table(mydata$TYPE)


```

(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2, fig.width = 12}

par(mfrow = c(1,2))

hist(mydata$VOLUME[mydata$TYPE=="I"], main="Volume for Infants",xlab="Infant Volume", xlim=c(0,1000), ylim=c(0,140), col="blue")
hist(mydata$VOLUME[mydata$TYPE=="ADULT"], main="Volume for Adults",xlab="Adult Volume", xlim=c(0,1000), ylim=c(0,140), col="red")

par(mfrow = c(1,1))

```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: The adult abalone volumes are closer to being normally distributed,while the infant abalone volumes are right skewed in distribution. he majority of the Infant volume is 300 or less whereas themajority of the adult volume lies between 100 and 700. While adult and infant abalone volumes have different means, there seem to be many difficulties separating infants from adults based on volume; it does not make sense to use volume as a separator between infant and adult abalones. ***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b, fig.width = 12, fig.height = 8}

L_SHUCK <- log10(mydata$SHUCK)
L_VOLUME <- log10(mydata$VOLUME)
mydata <- cbind(mydata, L_SHUCK, L_VOLUME)

par(mfrow = c(2,2))

plot(mydata$VOLUME, mydata$SHUCK, main = "SHUCK by VOLUME",col = mydata$CLASS, xlab = "VOLUME", ylab = "SHUCK",pch = 20)
legend(0,160,unique(mydata$CLASS),col = 1:length(mydata$CLASS), pch = 20,bty = "n", title = "CLASS")

plot(mydata$L_VOLUME, mydata$L_SHUCK, main = "Log10 SHUCK by Log10 VOLUME", col = mydata$CLASS, xlab = "L_VOLUME", ylab = "L_SHUCK",pch = 20)
legend(0.5,2.25,unique(mydata$CLASS),col = 1:length(mydata$CLASS), pch = 20,bty = "n", title = "CLASS")

plot(mydata$VOLUME, mydata$SHUCK, main = "SHUCK by VOLUME",col = mydata$TYPE, xlab = "VOLUME", ylab = "SHUCK",pch = 20)
legend(0,160,unique(mydata$TYPE),col = 1:length(mydata$TYPE), pch = 20,bty = "n", title = "TYPE")

plot(mydata$L_VOLUME, mydata$L_SHUCK, main = "Log 10 SHUCK by Log 10VOLUME", col = mydata$TYPE, xlab = "L_VOLUME", ylab = "L_SHUCK",pch = 20)
legend(0.5,2.25,unique(mydata$TYPE),col = 1:length(mydata$TYPE), pch = 20,bty = "n", title = "TYPE")

par(mfrow = c(1,1))

```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: Log transformation of the variables greatly reduced the variability in the plot. The general linear relationship does not change but the log transformation provides more clarity to the relationship. Additionally, most of the points are now clustered in the top right of the plot rather than the bottom left. The two different types are much easier to analyze in these plots, as infants have very clearly the lower volume and shuck than adults. We can determine that A3 and A4 have similar volumesbut A3 has higher shuck weight. We can note that A1 has the lowest shuckweight and volume. ***


***####   Section 4: (5 points) ####***

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. You are given code in (4)(a1) to reclassify the infants in classes A4 and A5 as ADULTS. 

```{r Part_4a1}

mydata$TYPE[mydata$CLASS == "A4" | mydata$CLASS == "A5"] <- "ADULT"
table(mydata$TYPE)

```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}

model <- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data = mydata)
summary(model)

```

**Essay Question:  Interpret the trend in CLASS level coefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: Classes A2 through A5 have negative coefficients, which means as they grow older, their shuck weight grows at a slower velocity while their volume velocity increases. This is consistent with the plots previously illustrated. Additionally, L_VOLUME has the largest predicting power due to its increasing impact on the outcomes. ***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: The coefficient of TYPE is -0.021 which is fairly significant., Other variables have more of an impact on the predictions,so its effect to making predictions is larger than that of classes, but not that significant upon L_SHUCK for the harvesting decision. In comparison, L_VOLUME has the biggest impact in predicting L_SHUCK for harvesting decisions. ***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

***#### Section 5: (5 points) ####***

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5am, fig.width = 12}

#histogram and QQ plot
par(mfrow = c(1,2))

hist(model$residuals, main = "Histogram of Residuals", col = "red",xlab = "Residuals")
qqnorm(model$residuals, main = "Q-Q Plot of Residuals", col = "blue")
qqline(model$residuals,col = "darkblue")

par(mfrow = c(1,1))

#skewness and kurtosis using rockchalk
skew_rock_r <- rockchalk::skewness(model$residuals)
kurt_rock_r_sub <- rockchalk::kurtosis(model$residuals)
kurt_rock_r <- rockchalk::kurtosis(model$residuals,excess = FALSE)

#skewness and kurtosis using momnents
skew_moments_r <- moments::skewness(model$residuals)
kurt_moments_r <- moments::kurtosis(model$residuals)

cat("Skewness using moments : ", skew_moments_r, "\n")
cat("Skewness using rockchalk : ", skew_rock_r, "\n")
cat("Kurtosis using moments : ", kurt_moments_r, "\n")
cat("Kurtosis using rockchalk after subtracting 3 : ", kurt_rock_r_sub, "\n")
cat("Kurtosis using rockchalk without subtracting : ", kurt_rock_r)



```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b, fig.width = 12, fig.height = 8}

#Plot the residuals versus L_VOLUME, coloring the data points by CLASS and TYPE 
par(mfrow = c(2,2))

plot(mydata$L_VOLUME, model$residuals, main = "Residuals by Log10 of Volume", col = mydata$CLASS, xlab = "L_VOLUME", ylab = "Residuals", cex =.8, xlim = c(0.5,4),pch = 20)
legend(3.5, 0.35, unique(mydata$CLASS), col =1:length(mydata$CLASS), pch=20, bty = "n", title = "CLASS", cex =.8)

plot(mydata$L_VOLUME, model$residuals, main = "Residuals by Log10 of Volume", col = mydata$TYPE, xlab = "L_VOLUME", ylab = "Residuals", cex =.8, xlim = c(0.5,4),pch = 20)
legend(3.5, 0.35, unique(mydata$TYPE), col =1:length(mydata$TYPE), pch=20, bty = "n", title = "TYPE", cex =.8)

boxplot(model$residuals ~ mydata$CLASS, main = "Residuals by CLASS", xlab = "Type", ylab= "Residuals",col = "purple")

boxplot(model$residuals ~ mydata$TYPE, main = "Residuals by TYPE", xlab = "Type", ylab ="Residuals",col = "orange")

par(mfrow = c(1,1))

#Test the homogeneity of variance
bartlett.test(model$residuals ~ mydata$CLASS)

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: Overall, we see that the residuals are evenly spread across y = 0. There is a slight concern about the spread seeming like it's in a cone shaped format, where there are more data points clustered to the right compared to the more spread out data on the left. L_VOLUME (and VOLUME itself) might be a useful parameter for harvesting decisions, but there could be a few more adjustments to make the residuals more evenly distributed. ***

-----

Harvest Strategy:

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.The Management needs to make a decision to implement 1 rule that meets the business goal.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

***#### Section 6: (5 points) ####***

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

```

(6)(b) Our first "rule" will be protection of all infants. We want to find a volume cutoff that protects all infants, but gives us the largest possible harvest of adults. We can achieve this by using the volume of the largest infant as our cutoff. You are given code below to identify the largest infant VOLUME and to return the proportion of adults harvested by using this cutoff. You will need to modify this latter code to return the proportion of infants harvested using this cutoff. Remember that we will harvest any individual with VOLUME greater than our cutoff.

```{r Part_6b}
# Largest infant volume
(max_inf_vol <- max(mydata$VOLUME[mydata$TYPE == "I"]))  # [1] 526.6383

# Proportion of adults harvested
cutoff_largest_a <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) /
  total.adults  # [1] 0.2476573

cutoff_largest_a

# Add code to calculate the proportion of infants harvested
cutoff_largest_i <- sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) /
  total.infants 

cutoff_largest_i

# If we use the largest infant volume, we harvest approximately 24.8% of adults and 0%,
# as expected, of infants.
sprintf("After using largest infant volume we harvest approximately %.1f percent of adults and %.1f percent of infants.",cutoff_largest_a*100,cutoff_largest_i*100)

```

(6)(c) Our next approaches will look at what happens when we use the median infant and adult harvest VOLUMEs. Using the median VOLUMEs as our cutoffs will give us (roughly) 50% harvests. We need to identify the median volumes and calculate the resulting infant and adult harvest proportions for both.

```{r Part_6c}
# Add code to determine the median infant volume:

med_inf_vol <- median(mydata$VOLUME[mydata$TYPE == "I"])
med_inf_vol

# Add code to calculate the proportion of infants harvested

cutoff_i_median_i <-sum(mydata$VOLUME[mydata$TYPE == "I"] > med_inf_vol)/total.infants
cutoff_i_median_i

# Add code to calculate the proportion of adults harvested

cutoff_a_median_i <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_inf_vol)/total.adults
cutoff_a_median_i

# If we use the median infant volume as our cutoff, we harvest almost 50% of our infants
# and a little more than 93% of our adults.

sprintf("After using median infant volume we harvest approximately %.1f percent of infants and %.1f percent of adults.",cutoff_i_median_i*100,cutoff_a_median_i*100)

# Add code to determine the median adult volume:

med_adult_vol <- median(mydata$VOLUME[mydata$TYPE == "ADULT"])
med_adult_vol

# Add code to calculate the proportion of infants harvested

cutoff_i_median_a <-sum(mydata$VOLUME[mydata$TYPE == "I"] > med_adult_vol)/total.infants
cutoff_i_median_a

# Add code to calculate the proportion of adults harvested

cutoff_a_median_a <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_adult_vol)/total.adults
cutoff_a_median_a

# If we use the median adult volume as our cutoff, we harvest almost 50% of adults
# and approximately 2.4% of infants.

sprintf("After using median adult volume we harvest approximately %.1f percent of adults and %.1f percent of infants.",cutoff_a_median_a*100,cutoff_i_median_a*100)

```

(6)(d) Next, we will create a plot showing the infant conserved proportions (i.e. "not harvested," the prop.infants vector) and the adult conserved proportions (i.e. prop.adults) as functions of volume.value. We will add vertical A-B lines and text annotations for the three (3) "rules" considered, thus far:  "protect all infants," "median infant" and "median adult." Your plot will have two (2) curves - one (1) representing infant and one (1) representing adult proportions as functions of volume.value - and three (3) A-B lines representing the cutoffs determined in (6)(b) and (6)(c).

```{r Part_6d, fig.width = 12, fig.height = 6}

plot(volume.value,prop.adults, xlab="", ylab="", type="l",xlim=c(0,1000), ylim=c(0,1), col="red")

par(new=T)
plot(volume.value,prop.infants, xlab="", ylab="", xaxt = "n",yaxt = "n",type="l", col="blue")

title(main="Proportion of Adults and Infants Protected", xlab="Volume",ylab="Proportion")
legend("bottomright", legend=c("Adult","Infant"), lty=1,col=c("red","blue"))

abline(h =.5, lty= "dashed")

abline(v=med_adult_vol, lty = "dashed")
text(med_adult_vol+30, .45, labels=round(med_adult_vol,1), cex=.8)

abline(v=med_inf_vol, lty = "dashed")
text(med_inf_vol+30, .45, labels=round(med_inf_vol,1), cex=.8)

abline(v=max_inf_vol, lty = "dashed")
text(max_inf_vol+30, .45, labels=round(max_inf_vol,1), cex=.8)

text(max_inf_vol+30, .40, "Max Infant Volume")
text(med_inf_vol+30, .40, "Median Infant Volume")
text(med_adult_vol+30, .40, "Median Adult Volume")

```

**Essay Question:  The two 50% "median" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: The median point at which the infants are split is much lower than the point at which the adults are split. This plot shows that infant abalones generally have smaller volumes than adult abalones as the 50%mark is reached at a much smaller volume. The infant curve is much steeper than the adult curve, and produces a volume significantly less in value than the adult abalones at the 50% mark. If 50% of the infant population needs to be protected, a volume threshold of at least134 must be met. If a volume cutoff of 384 is met than at least 50% of adult abalones would be harvested, with only a minimal amount of infants being harvested as well. ***

-----


More harvest strategies:

This part will address the determination of a cutoff volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. In other words, we want to find the volume value such that the vertical distance between the infant curve and the adult curve is maximum. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

***#### Section 7: (10 points)  ####***

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}

plot(volume.value, ((1-prop.adults) - (1-prop.infants)), main = "Difference VS Volume Value", ylab = "Difference in harvested proportions", xlab = "Volume", cex = 0.05, col = "steelblue")
abline(v=med_adult_vol, lty = "dashed")
abline(v=med_inf_vol, lty = "dashed")

```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

Include, side-by-side, the plot from (6)(d) but with a fourth vertical A-B line added. That line should intercept the x-axis at the "max difference" volume determined from the smoothed curve here.

```{r Part_7c, fig.width = 12, fig.height = 6}

propharvestdiff <- ((1 - prop.adults) - (1 - prop.infants)) 
head(propharvestdiff)

cutoff1 <- volume.value[which.max(smooth.difference)]
sprintf("Maximum differnce volume is %.2f",cutoff1)

par(mfrow = c(1,2))

plot(volume.value, propharvestdiff, type="l", main="Difference in Harvest Proportions", col="red", xlab="Volume", ylab="Difference")
lines(smooth.difference, type="l", col="orange")
abline(v=volume.value[which.max(smooth.difference)], col="blue",lty=2)

text(volume.value[which.max(smooth.difference)],propharvestdiff[which.max(smooth.difference)],label=round(volume.value[which.max(smooth.difference)],3), cex=.8, pos=1)
legend("topright", legend=c("Difference vs Volume","Smooth"), lty=1,col=c("red","orange"))

plot(volume.value,prop.adults, xlab="", ylab="", type="l",xlim=c(0,1000), ylim=c(0,1), col="green")
par(new=T)
plot(volume.value,prop.infants, xlab="", ylab="", type="l",  xaxt = "n",yaxt = "n", col="blue")

title(main="Proportion of Adults and Infants Protected", xlab="Volume",ylab="Proportion")
legend("bottomright", legend=c("Adult","Infant"), lty=1,col=c("green","blue"))

abline(h =.5, lty= "dashed")

abline(v=med_adult_vol, lty = "dashed")
text(med_adult_vol+30, .46, labels=round(med_adult_vol,1), cex=.8)

abline(v=med_inf_vol, lty = "dashed")
text(med_inf_vol+30, .26, labels=round(med_inf_vol,1), cex=.8)

abline(v=max_inf_vol, lty = "dashed")
text(max_inf_vol+30, .86, labels=round(max_inf_vol,1), cex=.8)

abline(v=volume.value[which.max(smooth.difference)], lty = "dashed")
text(volume.value[which.max(smooth.difference)]+30, .66, labels=round(volume.value[which.max(smooth.difference)],1), cex=.8)

text(max_inf_vol+30, .80, "Max Infant Volume")
text(volume.value[which.max(smooth.difference)]+30, .60, "Max Difference Volume")
text(med_inf_vol+30, .20, "Median Infant Volume")
text(med_adult_vol+30, .40, "Median Adult Volume")

```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions. We will actually calculate these proportions in two ways:  first, by 'indexing' and returning the appropriate element of the (1 - prop.adults) and (1 - prop.infants) vectors, and second, by simply counting the number of adults and infants with VOLUME greater than the volume threshold of interest.

Code for calculating the adult harvest proportion using both approaches is provided.

```{r Part_7d}

(1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332
# OR,
sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] >
      volume.value[which.max(smooth.difference)]) / total.adults # [1] 0.7416332

#calculating the infants harvest proportion
(1 - prop.infants)[which.max(smooth.difference)]


```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

***####  Section 8: (10 points)  ####***

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions. You may use either the 'indexing' or 'count' approach, or both.

```{r Part_8a}

volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786

cutoff2 <- volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &mydata$TYPE == "I", "VOLUME"])][1]

(1-prop.infants)[which(volume.value == cutoff2)]
(1-prop.adults)[which(volume.value == cutoff2)]

sprintf("When max difference cutoff %.4f is used: ",cutoff2)
sprintf("Adult  harvest proportions is: %.4f",(1-prop.adults)[which(volume.value == cutoff2)])
sprintf("Infant harvest proportion is: %.4f",(1-prop.infants)[which(volume.value == cutoff2)])

```

(8)(b) Next, append one (1) more vertical A-B line to our (6)(d) graph. This time, showing the "zero A1 infants" cutoff from (8)(a). This graph should now have five (5) A-B lines:  "protect all infants," "median infant," "median adult," "max difference" and "zero A1 infants."

```{r Part_8b, fig.width = 12}

plot(volume.value,prop.adults, xlab="", ylab="", type="l",xlim=c(0,1000), ylim=c(0,1), col="red")
par(new=T)
plot(volume.value,prop.infants, xlab="", ylab="", type="l", xaxt = "n",yaxt = "n", col="blue")

title(main="Proportion of Adults and Infants Protected", xlab="Volume",ylab="Proportion")
legend("bottomright", legend=c("Adult","Infant"), lty=1,col=c("red","blue"))

abline(h =.5, lty= "dashed")

abline(v=med_adult_vol, lty = "dashed")
text(med_adult_vol+20, .46, labels=round(med_adult_vol,1), cex=.8)

abline(v=med_inf_vol, lty = "dashed")
text(med_inf_vol+20, .26, labels=round(med_inf_vol,1), cex=.8)

abline(v=max_inf_vol, lty = "dashed")
text(max_inf_vol+20, .86, labels=round(max_inf_vol,1), cex=.8)

text(max_inf_vol+30, .80, "Max Infant Volume")
text(med_inf_vol+30, .20, "Median Infant Volume")
text(med_adult_vol+30, .40, "Median Adult Volume")

abline (v = cutoff2, lty = "dashed")
text(cutoff2+20, .65, "Zero A1 Inf")
text(cutoff2+10, .60, labels = round(cutoff2,1))

abline (v = cutoff1, lty = "dashed")
text(cutoff1+20, .75, "Max Diff")
text(cutoff1+10, .80, labels = round(cutoff1,1))
     


```


***#### Section 9: (5 points) ####***

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (6), (7) and (8) on this plot and label each. 

```{r Part_9, fig.width = 8.5}

prop_harvest_a <- 1 - prop.adults
prop_harvest_i <- 1 - prop.infants

plot(prop_harvest_i, prop_harvest_a, xlab = "1 - prop.infants", ylab = "1 - prop.adults", main = "ROC Curve ofAdult and Infant Harvest Proportions", type="l", col = "blue")

points(prop_harvest_i[which(volume.value==cutoff1)],prop_harvest_a[which(volume.value==cutoff1)],col = "red")
text(.3, .7, labels="Max difference vol=262.1", cex=.8)

points(prop_harvest_i[which(volume.value==cutoff2)],prop_harvest_a[which(volume.value==cutoff2)],col = "red")
text(.42, .82, labels="Zero A1 inf vol=206.8", cex=.8)

points(cutoff_i_median_a, cutoff_a_median_a,col = "red")
text(.15, .52, labels="Median Adult vol=384.6", cex=.8)

points(cutoff_i_median_i, cutoff_a_median_i,col = "red")
text(.6, .9, labels="Median inf vol=133.8", cex=.8)

points(cutoff_largest_i, cutoff_largest_a,col = "red")
text(.1, .2, labels="Protect All Infants vol=526.6", cex=.8)

abline(0,1, lty= "dashed",col = "purple")


```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}

area <- flux::auc(x = (1 - prop.infants), y = (1 - prop.adults))
cat("Area under ROC curve: ", area)

```


***#### Section 10: (10 points) ####***

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population

To calculate the total harvest proportions, you can use the 'count' approach, but ignoring TYPE; simply count the number of individuals (i.e. rows) with VOLUME greater than a given threshold and divide by the total number of individuals in our dataset.
 	
```{r Part_10} 	

table_harvest <- matrix(c(round(max_inf_vol,1), round(cutoff_largest_a,3), round(cutoff_largest_i,3), 
                          round(sum(mydata$VOLUME>max_inf_vol)/(total.adults+total.infants),3),
                          round(med_inf_vol,1), round(cutoff_a_median_i,3), round(cutoff_i_median_i,3), 
                          round(sum(mydata$VOLUME>med_inf_vol)/(total.adults+total.infants),3),
                          round(med_adult_vol,1), round(cutoff_a_median_a,3), round(cutoff_i_median_a,3), 
                          round(sum(mydata$VOLUME>med_adult_vol)/(total.adults+total.infants),3),
                          round(cutoff1,1),
                          round(prop_harvest_a[which.max(smooth.difference)],3),
                          round(prop_harvest_i[which.max(smooth.difference)],3),
                          round(sum(mydata$VOLUME>cutoff1)/(total.adults+total.infants),3),
                          round(cutoff2,1),
                          round(prop_harvest_a[which(volume.value==cutoff2)],3),
                          round(prop_harvest_i[which(volume.value==cutoff2)],3),
                          round(sum(mydata$VOLUME>cutoff2)/(total.adults+total.infants),3)), 
                          ncol=4, byrow=TRUE)

colnames(table_harvest) <- c("Volume",  "TPR", "FPR", "Total Harvest")
rownames(table_harvest) <- c("Protect all Infants", "Median Infant","Median Adult", "Max Difference", "Zero A1 Infants")

table_harvest_df <- data.frame(table_harvest)
print(table_harvest_df)

```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the five cutoffs determined in this assignment.**   

***Answer: The cutoff determined by the maximum difference between adult and infant abalone proportions harvested shows both a low proportion of both types harvested and the lowest true positive rate. The Zero A1 Infant cutoff has the highest false positive rate observed. The median adult cutoff produces the lowest false positive rate. The median infant cutoff produces the highest proportion of abalones harvested but the lowest volume. The Protect all infants cutoff yields the highest volume but the lowest proportion of abalones harvested.  ***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for  implementation of a cutoff?  
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer: I would make the recommendation to expand on differentiating and labeling different abalone. We did not seem to have an appropriate method to distinguish infant and adult abalones so this additional useful data could have been further integrated into this analysis. Finding a standard approach that could identify abalones would help reduce false positives. It is also important to note and I would encourage the implementation of sustainable practices particularly for the infant abalone population, which includes preventing the infant abalone population from being harvested at any cutoff. Altogether, more background information would have been useful as well as standardized testing approaches which should be further utilized in the future.  I would make the recommendation to use the max difference cutoff due to its relationship between volume and proportion harvested, to help with sustaining the population, however since protect all infants minimizes the FPR, I would urge the team to implement the protect all infants cutoff. ***